---
title: "Untitled"
output: html_document
date: "2025-09-11"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)
theme_set(theme_light())
```

Primary Questions:

What drives rating performance across different player segments?
How does chess talent distribution vary globally and demographically?
Can we identify patterns in player development and career trajectories?
What factors predict high-performance outliers?

Secondary Investigations:

Rating inflation/deflation trends over time
Federation-level competitive strength analysis
Gender participation and performance gaps
Title progression pathways and bottlenecks
Age-related performance curves




```{r}
library(tidyverse)

fide_ratings_august <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-09-23/fide_ratings_august.csv') %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(month = "August")

fide_ratings_september <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-09-23/fide_ratings_september.csv') %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(month = "September")

combined <- bind_rows(fide_ratings_august, fide_ratings_september)

head(combined)
```



```{r}
combined %>%
  group_by(sex) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    total_players = n()
  )

combined %>%
  group_by(title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    total_players = n()
  ) %>%
  arrange(desc(avg_rating))

combined %>%
  group_by(fed) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    total_players = n()
  ) %>%
  filter(total_players >= 500) %>%
  arrange(desc(avg_rating))

current_year <- as.numeric(format(Sys.Date(), "%Y"))

combined %>%
  filter(!is.na(bday)) %>%
  mutate(age = current_year - bday) %>%
  group_by(age) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    total_players = n()
  ) %>%
  filter(total_players >= 100) %>%
  arrange(age)
```



```{r}
combined_ratings <- inner_join(fide_ratings_august, fide_ratings_september, by = "id", suffix = c("_aug", "_sept")) %>%
  mutate(
    rating_change = rating_sept - rating_aug,
    rating_bin = cut(rating_aug,
                     breaks = seq(0, 3000, by = 200),
                     labels = paste0(seq(0, 2800, by = 200), "-", seq(200, 3000, by = 200)),
                     right = FALSE)
  )

combined_ratings %>%
  group_by(rating_bin) %>%
  summarise(
    avg_change = mean(rating_change, na.rm = TRUE),
    total_players = n()
  ) %>%
  filter(!is.na(rating_bin)) %>%
  ggplot(aes(x = rating_bin, y = avg_change)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Rating Change by Player Rating Bin",
    x = "August Rating Bin",
    y = "Average Rating Change (Aug to Sep)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
model <- lm(rating_change ~ rating_aug + sex_aug + title_aug + k_aug, data = combined_ratings)
summary(model)
```


